---
title: "Predicting Weight Lifting Exercise Quality"
author: "Hannu Kivimäki"
date: "2016-04-02"
output:
   html_document:
      toc: true
      toc_depth: 4
      theme: readable
---
<style type="text/css">
/* Reduce title font size. */
h1.title { font-size: 1.6em; }

/* Reduce pre block line height to normal. */
div pre {
  line-height: normal;
}

/* Enable page breaks for printing. */
@media all {
    .page-break	{ display: none; }
}
@media print {
    .page-break	{ display: block; page-break-before: always; }
    a[href]:after { content: none !important; }
}
</style>

### Synopsis
Activity trackers such as Nike FuelBand, Microsoft Band or Polar Loop are used
by people interested in measuring their overall physical activity and training
performance. However, while the amount of activity is quite easily measured,
the quality if often neglected.

In this report we present an example of how to use machine learning to predict how
well a basic weight lifting exercise was executed, based on sensor data from
six participants with accelerometers attached on the belt, forearm, arm and
dumbbell (the weight). We cover the prediction model building process, cross
validation steps and error rates. Finally, we use the model to classify
twenty test cases.

### Data Processing
```{r readcsv, echo = FALSE}
testing  <- read.csv("pml-testing.csv")
training <- read.csv("pml-training.csv")
```
The original weight lifting exercises dataset [1] was split into a training set
of `r nrow(training)` observations and a testing set of `r nrow(testing)`
observations. The data had `r ncol(training)` variables.

The first step in cleaning the data was to remove variables for which all the 
values were missing in the testing data set, as these would have been useless
for modelling. Additionally, columns 1-7 containing metadata such as name
of the person and timestamps were removed:

```{r validcols, echo = TRUE}
validCols <- colSums(is.na(testing)) != nrow(testing) # remove N/A columns
validCols[1:7] <- FALSE                               # remove names, timestamps etc.
training <- training[, validCols]
testing  <- testing[, validCols]
```

After pruning the data had `r ncol(training)` variables - measurements from
gyroscopes and accelerometers, with additional characteristics such minimum
and maximum values, standard deviations and variances. No values were missing,
i.e. no imputation was necessary.

For testing data, the last column (_problem\_id_) provides an ID for each case.
In the training data set, the last column (_classe_) describes how the
_Unilateral Dumbbell Biceps Curl_ weight exercise was actually performed:

 * Class A: exactly according to the specification
 * Class B: throwing the elbows to the front
 * Class C: lifting the dumbbell only halfway
 * Class D: lowering the dumbbell only halfway
 * Class E: throwing the hips to the front

Classes B-E represent common problems in the exercise. The goal is to use the
measurement data to recognize the correct class for each observed case.

### References
[1]  _Weight Lifting Exercises Dataset_ (http://groupware.les.inf.puc-rio.br/har)

