---
title: "Predicting Weight Lifting Exercise Performance"
author: "Hannu Kivimäki"
date: "2016-04-03"
output:
   html_document:
      toc: true
      toc_depth: 4
      theme: readable
---
<style type="text/css">
/* Reduce title font size. */
h1.title { font-size: 1.6em; }

/* Reduce pre block line height to normal. */
div pre {
  line-height: normal;
}

/* Enable page breaks for printing. */
@media all {
    .page-break	{ display: none; }
}
@media print {
    .page-break	{ display: block; page-break-before: always; }
    a[href]:after { content: none !important; }
}
</style>

```{r loadlibraries, echo = FALSE, message = FALSE, warning = FALSE}
library(caret)
library(randomForest)
library(knitr)
```

### Synopsis
Activity trackers such as Nike FuelBand, Microsoft Band or Polar Loop are used
by people interested in measuring their overall physical activity and training
performance. While the amount of activity is quite easily measured, the quality
is often neglected.

In this report we present an example of how to use machine learning to predict
how well a basic weight lifting exercise was executed, based on sensor data from
six participants with accelerometers attached on the belt, forearm, arm and
dumbbell (the weight). We cover the data pruning and prediction model fitting
processes and use the model to classify twenty test cases.

### Data Processing
For the purposes of this machine learning assignment, the original weight
lifting exercises data [1] was provided as two subsets:

```{r readcsv, echo = TRUE, cache = TRUE}
testing  <- read.csv("pml-testing.csv")
training <- read.csv("pml-training.csv")
```

The given training set had `r nrow(training)` observations and the testing set
`r nrow(testing)` observations. The data had `r ncol(training)` variables.

The first step in cleaning the data was to remove variables for which all the 
values were missing in the testing data set, as these would have been impossible
to use for predictions. Additionally, columns 1-7 containing metadata such as name
of the person and timestamps were removed:

```{r validcols, echo = TRUE, cache = TRUE}
validCols <- colSums(is.na(testing)) != nrow(testing) # remove N/A columns
validCols[1:7] <- FALSE                               # remove names, timestamps etc.
training <- training[, validCols]
testing  <- testing[, validCols]
```

After pruning the data had `r ncol(training)` variables left - measurements from
gyroscopes and accelerometers, with additional characteristics such minimum
and maximum values, standard deviations and variances. No values were missing
in either subsets, i.e. no imputation was needed.

For testing data, the last 53rd column (_problem\_id_) provides an ID for each
case. In the training data set, the last column (_classe_) describes how the
_Unilateral Dumbbell Biceps Curl_ weight exercise was actually performed:

 * Class A: exactly according to the specification
 * Class B: throwing the elbows to the front
 * Class C: lifting the dumbbell only halfway
 * Class D: lowering the dumbbell only halfway
 * Class E: throwing the hips to the front

Classes B-E represent common problems in the exercise. The classes are defined
in the paper [2] by Velloso et al. The goal is to use the sensor measurement
data to recognize the performance class for each observed case.

### Fitting a Model
Since the outcome variable is a factor with multiple categories (class A-E),
a tree-based approach is better than linear models. Furthermore, as the
training set is fairly large with thousands of observations, we can use
cross-validation for feature selection.

Random forests is an accurate machine learning technique for classification
tasks. However a model with all `r ncol(training) - 1` weight lifting
exercise variables would be very complex and possibly impractically slow
for making predictions on new data. To simplify the model, we use
cross-validation to narrow the selection of features to just the most
important ones:

```{r rfcv, echo = TRUE, cache = TRUE, results = 'hide'}
set.seed(1234)
rftest <- rfcv(training[, -53], training$classe, cv.fold = 10,
               scale = "log", step = 0.75)
rftest$error.cv
```

*rfcv()* starts with all predictors, splits the the training data using 10-fold
cross-validation and calculates the average prediction error rate. It then
preserves 75% of most important predictors (53 &rarr; 39) and starts the process
again. The process is repeated until there is just one predictor left.

The result shows that with just nine predictors the cross-validated classification
error rate stays below one percent (`r round(rftest$error.cv["9"], 3)`).
Adding more variables would have very little impact, other than causing
overfitting:

<div style="width:90%; padding-left: 50px;">
```{r rfcvtable, echo = FALSE}
kable(t(data.frame(rftest$error.cv)), digits = 3, row.names = F,
      caption = "Number of Predictors vs. Cross-validated Error Rate")
```

</div>

To find out which nine variables should be included in the model, we run the
random forest algorithm once with all `r ncol(training) - 1` predictors,
calculate the importance measure for each variable and choose top 9:

```{r importance, echo = TRUE, cache = TRUE}
fit52 <- train(classe ~ ., data = training, method = "rf")
top9  <- order(fit52$finalModel$importance, decreasing = T)[1:9]

```
<div style="width:90%; padding-left: 50px;">
```{r importancetable, echo = FALSE}
kable(fit52$finalModel$importance[top9,, drop = F], digits = 2, row.names = T,
      caption = "Top 9 Weigth Lifting Exercise Performance Predictors")
```

</div>

This leaves us the predictors to use in the final model.

### Final Model
After finding the most important predictors, we are ready to fit our final
random forests model:

```{r finalfit, echo = TRUE, cache = TRUE}
fmla <- as.formula(paste("classe ~ ",
                         paste(colnames(training)[top9], collapse = " + ")))
fit9 <- train(fmla, data = training, method = "rf")
```

Notice that we use the entire training set in the fitting process, without
diving the data further to training and validation subsets. This is intentional,
as _"in random forests, there is no need for cross-validation or a separate test
set to get an unbiased estimate of the test set error."_ [3].

```{r finalmodel, echo = TRUE}
fit9$finalModel

```
```{r ooberror, echo = FALSE}
# https://github.com/cran/randomForest/blob/master/R/print.randomForest.R
ooberr <- round(fit9$finalModel$err.rate[fit9$finalModel$ntree, "OOB"] * 100,
                digits = 2)
```

The reported out-of-bag (OOB) error rate `r ooberr`% is an estimate for our
model's out of sample error.

### Predicting Performance Classes
Once the final model has been built, it can be applied to predict the performance
class A-E for the the weight lifting exercise testing data set observations:
```{r predict, echo = TRUE}
predict(fit9, newdata = testing)
```


### References
[1]  _Weight Lifting Exercises Dataset_ (http://groupware.les.inf.puc-rio.br/har)

[2]  _Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. QualitativeActivity Recognition of Weight Lifting Exercises. Proceedings of 4th Augmented Human (AH) International Conference in cooperation with ACM SIGCHI (Augmented Human'13) . Stuttgart, Germany: ACM SIGCHI, 2013._  (http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201)

[3]  _Leo Breiman and Adele Cutler: Random Forests, The out-of-bag (oob) error estimate_ (http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr)
